Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-01-06 E-Hypergraphs-LICS24.tex
changed arxiv in bib wherever possible

Recalling the correspondence between DAG representations of algebraic terms and  string-diagrams for morphisms of a suitable Cartesian category
makes perspicuous the correspondence between (acyclic) e-graphs and morphisms of a suitably enriched Cartesian category, taken as string diagrams~\cite{noauthor_09083347_nodate,joyal_geometry_1991, mellies_functorial_2006}. We extend the usual vocabulary of string diagrams for symmetric monoidal categories with the additional generator  
\[
\infer{\phi_1 + \ldots + \phi_n: A \to B}{\{\phi_i: A \to B\}_{i \in \{1,\ldots,n\}}}
\]
The ability to take formal (non-empty) joins of morphisms is used to model the equivalence class structure of e-graphs.

@@

In this section,  we recall the fundamental definitions and results of string diagram rewrite theory for symmetric monoidal categories,  building to the recapitulation of the correspondence between string diagram rewriting and an appropriate notion of double pushout (DPO) rewriting of certain hypergraphs,  as established in~\cite{Frobenius, Frobenius2}.

@@

In ~\cite{Frobenius},  hypergraphs \textit{with interfaces} and their DPO rewriting are shown to be sound and complete for SMTs including a Frobenius algebra. 
Hypergraphs generalize graphs by allowing edges to have multiple sources and targets. 
When interpreting string diagrams as hypergraphs,  morphisms are interpreted as edges,  and wires as vertices.  
However,  there is also a need to specify which vertices are to be considered as \textit{input} and \textit{outputs},  corresponding to the dangling wires at the bottom and top of a string diagram. 
This gives rise to the definition of a hypergraph with interfaces.  
Subsequent work \cite{Frobenius2} builds on this correspondence to give a restriction of hypergraphs with interfaces (called \textit{monogamous} directed acyclic),  leading to a correspondence between symmetric monoidal string diagrams and the category of such hypergraphs.

@@

Following~\cite{Frobenius2} morphisms of $\textbf{PROP}(\Sigma)$ can be interpreted as particular cospans of discrete hypergraphs,  which we call hypergraphs with interfaces. 
\begin{definition}[Hypergraphs with interfaces]
\label{def:cspd}
Define the category of \emph{hypergraphs with interfaces} $\catname{HypI(\Sigma)}$ as the full sub-category of $\catname{Csp(Hyp(\Sigma))}$ with discrete hypergraphs as objects.
\end{definition}

@@

In order to represent only morphisms of $\textbf{PROP}(\Sigma)$,  it is necessary to restrict our attention to the following subset of hypergraphs with interfaces. 
\begin{definition}[Category of MDA Hypergraphs with Interfaces]
\label{def:monogamy_hyp}
We call a cospan $n \xrightarrow{f} \mathcal{G} \xleftarrow{g} m$ in $\HypI{\Sigma}$ \emph{monogamous directed acyclic (MDA)} if:
\begin{enumerate}
    \item $\mathcal{G}$ is a directed acyclic hypergraph;
    \item $f$ and $g$ are monomorphisms;
    \item the in-degree and out-degree of every vertex is at most $1$;
    \item vertices with in-degree $0$ are precisely the image of $f$; and
    \item vertices with out-degree $0$ are precisely the image of $g$.
\end{enumerate}
We denote by $\MdaCospans$ the subcategory of $\HypI{\Sigma}$ of MDA-cospans. 
\end{definition}
Note that $\MdaCospans$ remains a symmetric monoidal category with the same equipment as before.  In fact,  due to 
\cite{Frobenius2},  we have the following theorem which says the category $\MdaCospans$ is in fact \textit{equivalent} to $\textbf{PROP}(\Sigma)$. 
\begin{theorem}[Corollary 26~\cite{Frobenius2}]
    $\textbf{PROP}(\Sigma) \cong \MdaCospans$
\end{theorem}

@@

To express the rewriting of (hyper-)graphs with interfaces,  \textit{i.e.},  the ones that represent morphisms in $\HypI{\Sigma}$, the DPO formalism has been extended to double-pushout rewriting with \textit{interfaces}  (DPOI).
In the case of $ \catname{Hyp(\Sigma)}$,  we wish rewrite rules to be pairs of discrete cospans of hypergraphs with matching interfaces: $n \xrightarrow{} \mathcal L \xleftarrow{} m$, $n \xrightarrow{} \mathcal R \xleftarrow{} m$.  By noting that a cospan $0 \xrightarrow{} \mathcal L \xleftarrow{[f,g]} n + m$ encodes the same data as a cospan $n \xrightarrow{f} \mathcal L \xleftarrow{g} m$ (Remark 3.14~\cite{Frobenius}) we can formulate a rewrite rule as a single span $\mathcal L \xleftarrow{} n+m \xrightarrow{} \mathcal R$,  and we can do similarly for the discrete cospans giving the interface to $\mathcal G$ and $\mathcal H$.  Thus we reach the following variant of DPO rewriting. 
\begin{definition}[DPOI rewriting]\label{def:dpoi}
Given a span of morphisms $\mathcal G \xleftarrow{} n+m \xrightarrow{} \mathcal H$ in $\textbf{Hyp}(\Sigma)$,  we say \textit{$\mathcal G$ rewrites to $\mathcal H$ with interface $n+m$ via rewrite rule $\mathcal L \xleftarrow{} i+j \xrightarrow{} \mathcal R$} if there exists an object $\mathcal C$ and morphisms which complete the following commutative diagram such that the two marked squares are pushouts.
\[
\begin{tikzcd}
\mathcal L \arrow[d, "m"'] & i+j \arrow[d] \arrow[l] \arrow[r] & \mathcal R \arrow[d]   \\
\mathcal{G}^{\urcorner}     & \mathcal C \arrow[l] \arrow[r]                      & ^{\ulcorner}\mathcal H \\
                  & n+m \arrow[lu] \arrow[u] \arrow[ru]          &              
\end{tikzcd}
\]
\end{definition}
We need to restrict the notions of pushout complement and of match in order to preserve monogamy and directed acylcicity.

@@

CHRIS: WANT TO STATE THE CORRESPONDENCE THEOREM WITH SMTS HERE? NEED TO DEFINE $\lceil-\rceil$ AND INTERPRETATION? Hmm
\begin{theorem}[Theorem 35~\cite{Frobenius2}]
For any $f,g \in \mathbf{SMT}(\Sigma, \mathcal{E})$,  we have that 
\[
	f = g \iff \llangle \lceil f \rceil \rrangle \Rrightarrow^*_{\mathfrak{R}} \llangle \lceil g \rceil \rrangle~, 
\]
where $\mathfrak{R} = \llangle \lceil \mathcal{E} \rceil \rrangle$ and $ \Rrightarrow^*_{\mathfrak{R}}$ denotes the existence of a sequence of convex rewrites. 
\end{theorem}
\section{E-Hypergraphs}

@@

Unfolding the definitions,  we can spell out the previous theorem in terms of DPO-rewriting. 
\begin{corollary}
For any $\phi, \psi \in \textbf{PROP}^{+}(\Sigma, \mathcal{E})$,  we have
\[
	\phi = \psi \quad \iff \quad \llbracket\phi \rrbracket \Rrightarrow_{\mathcal{S}, \mathcal{E}} \llbracket\psi\rrbracket~ . 
\]
 \end{corollary}
CHRIS: INCORPORATING EXTRA EQUATIONS OUGHT TO BE IMMEDIATE? 
%
%\begin{theorem}[Theorem 35~\cite{Frobenius2}]
%
%    $a \Rightarrow_{R} b$ iff $\llangle {}^{\ulcorner} a {}^{\urcorner} \rrangle \Rrightarrow_{\llangle {}^\ulcorner R {}^\urcorner \rrangle} \llangle {}^{\ulcorner}b {}^{\urcorner} \rrangle$.
%    Where $\llangle \cdot \rrangle : \textsf{PROP}(\Sigma) + \textsf{Frob} \to \catname{Csp_{D}(Hyp_{\Sigma})}$ and $\Rrightarrow$ is a convex rewriting relation for morphisms in $\catname{Hyp_{\Sigma}}$.
%\end{theorem}
%
%\begin{proposition}
%    $\llangle {}^{\ulcorner} a {}^{\urcorner} \rrangle \Rrightarrow_{\llangle {}^\ulcorner R {}^\urcorner \rrangle} \llangle {}^{\ulcorner}b {}^{\urcorner} \rrangle$ iff $[\mathcal{J}(a)] \Rrightarrow_{[R]}^{+} [\mathcal{J}(b)]$.
%     Note that $a$ and $b$ are from $\textsf{PROP}(\Sigma)$.
%\end{proposition}
%
%\begin{theorem}
%    $a \Rightarrow_{\langle l, r \rangle}^{+} b$ in $\textsf{PROP}^{+}$ iff $[a] \Rrightarrow_{\langle [l],[r] \rangle} [b]$ in $\sfrac{\WellTypedMdaEcospans}{\mathcal{S}}$.
%\end{theorem}

@@

\begin{remark}
    Using functional notation we can reformulate the notion of homomorphism of two e-hypergraphs in the following way.
    \begin{definition}{E-hypergraphs homomorphism}
        \label{def:e-homo-2}    
        A homomorphism $\phi: \mathcal{F} \to \mathcal{G}$ of e-hypergraphs $\mathcal{F},\mathcal{G}$ is a pair of functions $\phi_V : V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E : E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that

@@

\begin{theorem}[Existence of pushouts in $\catname{EHyp_{\Sigma}}$]
\label{th:existence_of_pushouts}
Consider the following span in $\catname{EHyp_{\Sigma}}$
% https://q.uiver.app/#q=WzAsMyxbMCwwLCJaIl0sWzIsMCwiWCJdLFswLDIsIlkiXSxbMCwxLCJmIl0sWzAsMiwiZyIsMl1d
\[\begin{tikzcd}
	Z && X \\
	\\
	Y
	\arrow["f", from=1-1, to=1-3]
	\arrow["g"', from=1-1, to=3-1]
\end{tikzcd}\]
such that
\begin{enumerate}
\label{pushout:assumptions}
    \item $Z$ is a \textit{descrete} e-hypergraph.
    \item $f,g$ are monos.
    \item $[f_{V}(v_i)) = [f_{V}(v_j))$ and $[g_{V}(v_i)) = [g_{V}(v_j))$ for all $v_{i},v_{j}$ in $V_{Z}$.
    \item If $[f_{V}(v)) \not = \varnothing$ then $[g_{V}(v)) = \varnothing$ and if $[g_{V}(v)) \not = \varnothing$ then $[f_{V}(v)) = \varnothing$.
    \item $\consistency^{\mu}(f(v_i)) = \consistency^{\mu}(f(v_j))$ and $\consistency^{\mu}(g(v_i)) = \consistency^{\mu}(g(v_j))$ for all $v_i,v_j$ in $V_{Z}$.
\end{enumerate}    
then the pushout $X +_{f,g} Y$ exists.
\end{theorem}
\begin{proof}
    Consider the diagram below.
    \[
        \begin{tikzcd}
        Z \arrow[r, "f"] \arrow[d, "g"']                                   & X \arrow[d, "\sfrac{\iota_1}{\sim R}"] \arrow[rdd, "j_1", bend left] &   \\
        Y \arrow[r, "\sfrac{\iota_2}{\sim R}"] \arrow[rrd, "j_2"', bend right] & \sfrac{X+Y}{\sim R} \arrow[rd, "u"]                              &   \\
                                                                        &                                                                  & Q
        \end{tikzcd}
    \]
    Then, the pushout of e-hypergraphs $X$ and $Y$ is computed in two steps.
    First, a coproduct of $X+Y$ is computed which is 
    \[
        X + Y = \{V_{X} + V_{Y}, E_{X} + E_{Y}, s_{X+Y}, t_{X+Y}, \consistency_{X+Y}, <_{c,X+Y} \}
    \]
    where $s_{X+Y} : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ which can be defined as a copairing $[s'_{X}, s'_{Y}] : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ and $s'_{X} : E_{X} \to (V_{X} + V_{Y})^{*}$, $s'_{Y} : E_{Y} \to (V_{X} + V_{Y})^{*}$ defined as $s'_{X} = s_{X};\iota_{1,V}^{*}$ and $s'_{Y} = s_{Y};\iota_{2,V}^{*}$ where $\iota_1,\iota_2$ are corresponding coproduct injections, similarly for $t_{X+Y}, <_{c,X+Y}, \consistency_{X+Y}$.
    We will omit labels as they are irrelevant to pushout construction.
    Consider relations
    \begin{align*}    
    S_{V} &= \{\\
          &(x_i,y_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; | \\
          &\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z) \text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }\\
          &\}
    \end{align*}
    \begin{align*}
    S_{E} &= \{\\
    &(x_i,y_j) \in (E_{X} + E_{Y}) \times (E_{X} + E_{Y})\; |\\
    &\exists z \in E_{Z} \; . \; x_i = f_{E};\iota_{1,E}(z) \text{ and } y_j = g_{E};\iota_{2,E}(z) \text{ where $x_i \in E_{X}$ and $y_j \in E_{Y}$ }
    &\}
    \end{align*}
and let relations $R_{V}$ and $R_{E}$ be their reflexive symmetric and transitive closures respectively.

@@

So far, the construction matches the construction of the pushout in $\catname{Hyp_{\Sigma}}$ which can be found in~\cite{Frobenius} and hence $s_{\sfrac{X+Y}{\sim}}$ and $t_{\sfrac{X+Y}{\sim}}$ are well-defined and $[-]$ is a homomorphism with respect to sources and targets.

@@

\begin{definition}[Definition 3.16~\cite{Frobenius}]
Morphisms $K \xrightarrow{i} L \xrightarrow{m} G$ satisfy the \textit{no-dangling} condition if, for every hyperedge not in the image of $m$, every node of its source and target is either (i) not in the image of $m$ or (ii) in the image of $i ; m$.
They satisfy the no-identification condition if any two nodes merged by $m$ are in the image of $i$.
\end{definition}

>>>>

Recalling the correspondence between DAG representations of algebraic terms and  string-diagrams for morphisms of a suitable Cartesian category
makes perspicuous the correspondence between (acyclic) e-graphs and morphisms of a suitably enriched Cartesian category, taken as string diagrams~\cite{Selinger_2010,joyal_geometry_1991, mellies_functorial_2006}. We extend the usual vocabulary of string diagrams for symmetric monoidal categories with the additional generator  
\[
\infer{\phi_1 + \ldots + \phi_n: A \to B}{\{\phi_i: A \to B\}_{i \in \{1,\ldots,n\}}}
\]
The ability to take formal (non-empty) joins of morphisms is used to model the equivalence class structure of e-graphs.

@@

In this section,  we recall the fundamental definitions and results of string diagram rewrite theory for symmetric monoidal categories,  building to the recapitulation of the correspondence between string diagram rewriting and an appropriate notion of double pushout (DPO) rewriting of certain hypergraphs,  as established in~\cite{bonchi_string_2022-1, bonchi_string_2022-2}.

@@

In ~\cite{bonchi_string_2022-1},  hypergraphs \textit{with interfaces} and their DPO rewriting are shown to be sound and complete for SMTs including a Frobenius algebra. 
Hypergraphs generalize graphs by allowing edges to have multiple sources and targets. 
When interpreting string diagrams as hypergraphs,  morphisms are interpreted as edges,  and wires as vertices.  
However,  there is also a need to specify which vertices are to be considered as \textit{input} and \textit{outputs},  corresponding to the dangling wires at the bottom and top of a string diagram. 
This gives rise to the definition of a hypergraph with interfaces.  
Subsequent work \cite{bonchi_string_2022-2} builds on this correspondence to give a restriction of hypergraphs with interfaces (called \textit{monogamous} directed acyclic),  leading to a correspondence between symmetric monoidal string diagrams and the category of such hypergraphs.

@@

Following~\cite{bonchi_string_2022-2} morphisms of $\textbf{PROP}(\Sigma)$ can be interpreted as particular cospans of discrete hypergraphs,  which we call hypergraphs with interfaces. 
\begin{definition}[Hypergraphs with interfaces]
\label{def:cspd}
Define the category of \emph{hypergraphs with interfaces} $\catname{HypI(\Sigma)}$ as the full sub-category of $\catname{Csp(Hyp(\Sigma))}$ with discrete hypergraphs as objects.
\end{definition}

@@

In order to represent only morphisms of $\textbf{PROP}(\Sigma)$,  it is necessary to restrict our attention to the following subset of hypergraphs with interfaces. 
\begin{definition}[Category of MDA Hypergraphs with Interfaces]
\label{def:monogamy_hyp}
We call a cospan $n \xrightarrow{f} \mathcal{G} \xleftarrow{g} m$ in $\HypI{\Sigma}$ \emph{monogamous directed acyclic (MDA)} if:
\begin{enumerate}
    \item $\mathcal{G}$ is a directed acyclic hypergraph;
    \item $f$ and $g$ are monomorphisms;
    \item the in-degree and out-degree of every vertex is at most $1$;
    \item vertices with in-degree $0$ are precisely the image of $f$; and
    \item vertices with out-degree $0$ are precisely the image of $g$.
\end{enumerate}
We denote by $\MdaCospans$ the subcategory of $\HypI{\Sigma}$ of MDA-cospans. 
\end{definition}
Note that $\MdaCospans$ remains a symmetric monoidal category with the same equipment as before.  In fact,  due to 
\cite{bonchi_string_2022-2},  we have the following theorem which says the category $\MdaCospans$ is in fact \textit{equivalent} to $\textbf{PROP}(\Sigma)$. 
\begin{theorem}[Corollary 26~\cite{bonchi_string_2022-2}]
    $\textbf{PROP}(\Sigma) \cong \MdaCospans$
\end{theorem}

@@

To express the rewriting of (hyper-)graphs with interfaces,  \textit{i.e.},  the ones that represent morphisms in $\HypI{\Sigma}$, the DPO formalism has been extended to double-pushout rewriting with \textit{interfaces}  (DPOI).
In the case of $ \catname{Hyp(\Sigma)}$,  we wish rewrite rules to be pairs of discrete cospans of hypergraphs with matching interfaces: $n \xrightarrow{} \mathcal L \xleftarrow{} m$, $n \xrightarrow{} \mathcal R \xleftarrow{} m$.  By noting that a cospan $0 \xrightarrow{} \mathcal L \xleftarrow{[f,g]} n + m$ encodes the same data as a cospan $n \xrightarrow{f} \mathcal L \xleftarrow{g} m$ (Remark 3.14~\cite{bonchi_string_2022-1}) we can formulate a rewrite rule as a single span $\mathcal L \xleftarrow{} n+m \xrightarrow{} \mathcal R$,  and we can do similarly for the discrete cospans giving the interface to $\mathcal G$ and $\mathcal H$.  Thus we reach the following variant of DPO rewriting. 
\begin{definition}[DPOI rewriting]\label{def:dpoi}
Given a span of morphisms $\mathcal G \xleftarrow{} n+m \xrightarrow{} \mathcal H$ in $\textbf{Hyp}(\Sigma)$,  we say \textit{$\mathcal G$ rewrites to $\mathcal H$ with interface $n+m$ via rewrite rule $\mathcal L \xleftarrow{} i+j \xrightarrow{} \mathcal R$} if there exists an object $\mathcal C$ and morphisms which complete the following commutative diagram such that the two marked squares are pushouts.
\[
\begin{tikzcd}
\mathcal L \arrow[d, "m"'] & i+j \arrow[d] \arrow[l] \arrow[r] & \mathcal R \arrow[d]   \\
\mathcal{G}^{\urcorner}     & \mathcal C \arrow[l] \arrow[r]                      & ^{\ulcorner}\mathcal H \\
                  & n+m \arrow[lu] \arrow[u] \arrow[ru]          &              
\end{tikzcd}
\]
\end{definition}
We need to restrict the notions of pushout complement and of match in order to preserve monogamy and directed acylcicity.

@@

CHRIS: WANT TO STATE THE CORRESPONDENCE THEOREM WITH SMTS HERE? NEED TO DEFINE $\lceil-\rceil$ AND INTERPRETATION? Hmm
\begin{theorem}[Theorem 35~\cite{bonchi_string_2022-2}]
For any $f,g \in \mathbf{SMT}(\Sigma, \mathcal{E})$,  we have that 
\[
	f = g \iff \llangle \lceil f \rceil \rrangle \Rrightarrow^*_{\mathfrak{R}} \llangle \lceil g \rceil \rrangle~, 
\]
where $\mathfrak{R} = \llangle \lceil \mathcal{E} \rceil \rrangle$ and $ \Rrightarrow^*_{\mathfrak{R}}$ denotes the existence of a sequence of convex rewrites. 
\end{theorem}
\section{E-Hypergraphs}

@@

Unfolding the definitions,  we can spell out the previous theorem in terms of DPO-rewriting. 
\begin{corollary}
For any $\phi, \psi \in \textbf{PROP}^{+}(\Sigma, \mathcal{E})$,  we have
\[
	\phi = \psi \quad \iff \quad \llbracket\phi \rrbracket \Rrightarrow_{\mathcal{S}, \mathcal{E}} \llbracket\psi\rrbracket~ . 
\]
 \end{corollary}
CHRIS: INCORPORATING EXTRA EQUATIONS OUGHT TO BE IMMEDIATE? 
%
%\begin{theorem}[Theorem 35~\cite{bonchi_string_2022-2}]
%
%    $a \Rightarrow_{R} b$ iff $\llangle {}^{\ulcorner} a {}^{\urcorner} \rrangle \Rrightarrow_{\llangle {}^\ulcorner R {}^\urcorner \rrangle} \llangle {}^{\ulcorner}b {}^{\urcorner} \rrangle$.
%    Where $\llangle \cdot \rrangle : \textsf{PROP}(\Sigma) + \textsf{Frob} \to \catname{Csp_{D}(Hyp_{\Sigma})}$ and $\Rrightarrow$ is a convex rewriting relation for morphisms in $\catname{Hyp_{\Sigma}}$.
%\end{theorem}
%
%\begin{proposition}
%    $\llangle {}^{\ulcorner} a {}^{\urcorner} \rrangle \Rrightarrow_{\llangle {}^\ulcorner R {}^\urcorner \rrangle} \llangle {}^{\ulcorner}b {}^{\urcorner} \rrangle$ iff $[\mathcal{J}(a)] \Rrightarrow_{[R]}^{+} [\mathcal{J}(b)]$.
%     Note that $a$ and $b$ are from $\textsf{PROP}(\Sigma)$.
%\end{proposition}
%
%\begin{theorem}
%    $a \Rightarrow_{\langle l, r \rangle}^{+} b$ in $\textsf{PROP}^{+}$ iff $[a] \Rrightarrow_{\langle [l],[r] \rangle} [b]$ in $\sfrac{\WellTypedMdaEcospans}{\mathcal{S}}$.
%\end{theorem}

@@

\begin{remark}
    Using functional notation we can reformulate the notion of a homomorphism between two e-hypergraphs in the following way.
    \begin{definition}
        \label{def:e-homo-2}    
        A \emph{homomorphism} $\phi: \mathcal{F} \to \mathcal{G}$ of e-hypergraphs $\mathcal{F},\mathcal{G}$ is a pair of functions $\phi_V : V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E : E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that

@@

\begin{theorem}[Existence of pushouts in $\catname{EHyp_{\Sigma}}$]
\label{th:existence_of_pushouts}
Consider the following span in $\catname{EHyp_{\Sigma}}$
% https://q.uiver.app/#q=WzAsMyxbMCwwLCJaIl0sWzIsMCwiWCJdLFswLDIsIlkiXSxbMCwxLCJmIl0sWzAsMiwiZyIsMl1d
\[\begin{tikzcd}
	Z && X \\
	\\
	Y
	\arrow["f", from=1-1, to=1-3]
	\arrow["g"', from=1-1, to=3-1]
\end{tikzcd}\]
such that
\begin{enumerate}
\label{pushout:assumptions}
    \item $Z$ is a \textit{descrete} e-hypergraph.
    \item $f,g$ are monos.
    \item $[f_{V}(v_i)) = [f_{V}(v_j))$ and $[g_{V}(v_i)) = [g_{V}(v_j))$ for all $v_{i},v_{j}$ in $V_{Z}$.
    \item If $[f_{V}(v)) \not = \varnothing$ then $[g_{V}(v)) = \varnothing$ and if $[g_{V}(v)) \not = \varnothing$ then $[f_{V}(v)) = \varnothing$.
    \item $\consistency^{\mu}(f(v_i)) = \consistency^{\mu}(f(v_j))$ and $\consistency^{\mu}(g(v_i)) = \consistency^{\mu}(g(v_j))$ for all $v_i,v_j$ in $V_{Z}$.
\end{enumerate}    
then the pushout $X +_{f,g} Y$ exists.
\end{theorem}
\begin{proof}
    Consider the diagram below.
    \[
        \begin{tikzcd}
        Z \arrow[r, "f"] \arrow[d, "g"']                                   & X \arrow[d, "\sfrac{\iota_1}{\sim R}"] \arrow[rdd, "j_1", bend left] &   \\
        Y \arrow[r, "\sfrac{\iota_2}{\sim R}"] \arrow[rrd, "j_2"', bend right] & \sfrac{X+Y}{\sim R} \arrow[rd, "u"]                              &   \\
                                                                        &                                                                  & Q
        \end{tikzcd}
    \]
    Then, the pushout of e-hypergraphs $X$ and $Y$ is computed in two steps.
    First, a coproduct of $X+Y$ is computed which is 
    \[
        X + Y = \{V_{X} + V_{Y}, E_{X} + E_{Y}, s_{X+Y}, t_{X+Y}, \consistency_{X+Y}, <_{c,X+Y} \}
    \]
    where $s_{X+Y} : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ which can be defined as a copairing $[s'_{X}, s'_{Y}] : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ and $s'_{X} : E_{X} \to (V_{X} + V_{Y})^{*}$, $s'_{Y} : E_{Y} \to (V_{X} + V_{Y})^{*}$ defined as $s'_{X} = s_{X};\iota_{1,V}^{*}$ and $s'_{Y} = s_{Y};\iota_{2,V}^{*}$ where $\iota_1,\iota_2$ are corresponding coproduct injections, similarly for $t_{X+Y}, <_{c,X+Y}, \consistency_{X+Y}$.
    We will omit labels as they are irrelevant to pushout construction.
    Consider relations
    \begin{align*}    
    S_{V} &= \{\\
          &(x_i,y_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; | \\
          &\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z)\\
          &\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }\\
          &\}
    \end{align*}
    \begin{align*}
    S_{E} &= \{\\
    &(x_i,y_j) \in (E_{X} + E_{Y}) \times (E_{X} + E_{Y})\; |\\
    &\exists z \in E_{Z} \; . \; x_i = f_{E};\iota_{1,E}(z) \text{ and } y_j = g_{E};\iota_{2,E}(z)\\
    &\text{ where $x_i \in E_{X}$ and $y_j \in E_{Y}$}\\
    &\}
    \end{align*}
and let relations $R_{V}$ and $R_{E}$ be their reflexive symmetric and transitive closures respectively.

@@

So far, the construction matches the construction of the pushout in $\catname{Hyp_{\Sigma}}$ which can be found in~\cite{bonchi_string_2022-1} and hence $s_{\sfrac{X+Y}{\sim}}$ and $t_{\sfrac{X+Y}{\sim}}$ are well-defined and $[-]$ is a homomorphism with respect to sources and targets.

@@

\begin{definition}[Definition 3.16~\cite{bonchi_string_2022-1}]
Morphisms $K \xrightarrow{i} L \xrightarrow{m} G$ satisfy the \textit{no-dangling} condition if, for every hyperedge not in the image of $m$, every node of its source and target is either (i) not in the image of $m$ or (ii) in the image of $i ; m$.
They satisfy the no-identification condition if any two nodes merged by $m$ are in the image of $i$.
\end{definition}
